Nodal Discontinuous Galerkin Method for the Euler Equations in GR

### 2018-09-24 Documents/Euler/dgEulerGR.tex
Changed style slightly and updated references and definitions files

\subsubsection{Time-Step Derivation/CFL Condition}
For the first-order DG method using forward-Euler time-stepping, we evolve the vector of conserved variables as:
\begin{equation}\label{Eq:1stOrderDG}
    \ol{\bU}^{n+1}_{i}=\ol{\bU}^{n}_{i}-\eta_{i}\left[\hat{\bF}\left(\ol{\bU}^{n}_{i},\ol{\bU}^{n}_{i+1}\right)-\hat{\bF}\left(\ol{\bU}^{n}_{i-1},\ol{\bU}^{n}_{i}\right)\right],
\end{equation}
where
\begin{equation}
    \ol{\bU}_{i}\equiv\f{1}{\Delta x_{i}}\int_{x_{\imh}}^{x_{\iph}}\bU_{i}\,dx,
\end{equation}
$\eta_{i}\equiv\Delta t_{i}/\Delta x_{i}$, and $\hat{\bF}$ is the numerical flux. In this document we use the local Lax-Friedrichs flux, defined as:
\begin{equation}\label{Eq:LLF}
    \hat{\bF}\left(a,b\right)=\f{1}{2}\left[\bF\left(a\right)+\bF\left(b\right) - \alpha_{ab}\left(b-a\right)\right],
\end{equation}
where $a$ and $b$ represent the state of the fluid in two different elements, $\alpha_{ab}$ is an estimate for the wave-speed:
\begin{equation}
    \alpha_{ab}=\text{max}\left[\alpha\left(a\right),\alpha\left(b\right)\right],
\end{equation}
and $\alpha$ is the largest (in absolute value) eigenvalue of the flux-Jacobian:
\begin{equation}
    \alpha=\left|\left|\f{\p\bF}{\p\bU}\right|\right|.
\end{equation}
Using this we define the following variables:
\begin{equation}\label{Eq:EigVals}
    \alpha_{\iph}=\text{max}\left[\alpha\left(\ol{\bU}_{i}\right),\alpha\left(\ol{\bU}_{i+1}\right)\right],\hspace{3em} \alpha_{\imh}=\text{max}\left[\alpha\left(\ol{\bU}_{i-1}\right),\alpha\left(\ol{\bU}_{i}\right)\right].
\end{equation}

@@

\subsubsection{Set of Admissible States}
We again consider a one-dimensional system of conservation laws, but this time with a curvilinear metric:
\begin{equation}
    \pd{\left(\sqrtgm\,\bU\right)}{t}+\pd{\left(\sqrtgm\,\bF\right)}{1}=\sqrtgm\,\bQ,
\end{equation}
where $\bU$ is given by:
\begin{equation}
    \bU\longrightarrow\begin{pmatrix} D\\ S_{1}\\ \tau\end{pmatrix}=\begin{pmatrix} \rho\,W\\ \rho\,h\,W^{2}\,v_{1}\\ \rho\,W\left(h\,W-1\right)-p\end{pmatrix}=\begin{pmatrix} \rho\,W\\ \rho\,h\,W^{2}\,\gamma_{1j}\,v^{j}\\ \rho\,W\left(h\,W-1\right)-p\end{pmatrix},
\end{equation}
$\bF\left(\bU\right)$ are the fluxes of those conserved quantities:
\begin{equation}
    \bF\left(\bU\right)\longrightarrow\begin{pmatrix}D\,v^{1}\\ S^{1}\,v_{1}+p\,\delta^{1}_{~1}\\ S^{1}-D\,v^{1}\end{pmatrix}=\begin{pmatrix}\rho\,W\,v^{1}\\ \rho\,h\,W^{2}\,v^{1}\,v_{1}+p\\\rho\,h\,W^{2}\,v^{1}-D\,v^{1}\end{pmatrix}=\begin{pmatrix}\rho\,W\,v^{1}\\ \rho\,h\,W^{2}\,\gamma_{1j}\,v^{1}\,v^{j}+p\\\rho\,h\,W^{2}\,v^{1}-D\,v^{1}\end{pmatrix},
\end{equation}
and $\bQ$ is a source term:
\begin{align}
    \bQ\longrightarrow\begin{pmatrix}0\\\f{1}{2}\,P^{kj}\,\p_{1}\gamma_{kj}\\0\end{pmatrix}&=\begin{pmatrix}0\\ \f{1}{2}\left[P^{11}\,\p_{1}\,\gamma_{11}+P^{22}\,\p_{1}\,\gamma_{22}+P^{33}\,\p_{1}\,\gamma_{33}\right] \\0\end{pmatrix}\\
    &=\begin{pmatrix}0\\ P^{11}\,h_{1}\,\pd{\,h_{1}}{1}+P^{22}\,h_{2}\,\pd{\,h_{2}}{1}+P^{33}\,h_{3}\,\pd{\,h_{3}}{1} \\0\end{pmatrix},
\end{align}
where we have used the fact that $\gamma_{jj}=\left(h_{j}\right)^{2}$. The $P^{kj}$ are components of the pressure tensor:
\begin{equation}
    P^{kj}=S^{k}\,v^{j}+p\,\gamma^{kj}=\gamma^{kj}\left[S_{a}\,v^{a}+p\right].
\end{equation}

@@

It is clear that the first requirement for $\bH_{2}$ is met, i.e. $\ol{D}>0$. The second requirement is:
\begin{align}
\ol{D}+\ol{\tau}&>\sqrt{\ol{D}^{2}+\left[\ol{S}_{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}\right] \left[\ol{S}^{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\p^{1}\,\gamma_{kk}}\right] }\\
&=\sqrt{\ol{D}^{2}+\left[\ol{S}_{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}\right] \left[\ol{S}^{1}+\gamma^{11}\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\p_{1}\,\gamma_{kk}}\right] }\\
&=\sqrt{\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}+\gamma^{11}\left[\f{\ol{P^{kk}\,\p_{1}\,\gamma_{kk}}}{2\left(1-\ve\right)}\right]^{2}\left(\Delta t\right)^{2}+\f{\ol{P^{kk}\,\p_{1}\,\gamma_{kk}}}{2\left(1-\ve\right)}\left[\ol{S}_{1}\,\gamma^{11}+\ol{S}^{1}\right]\Delta t}\\
&=\sqrt{a\left(\Delta t\right)^{2}+b\,\Delta t+c}\geq0,
\end{align}
where
\begin{align}
    a&=\left[\f{\ol{P^{kk}\,\p_{1}\,\gamma_{kk}}}{2\,\sqrt{\ol{\gamma_{11}}}\left(1-\ve\right)}\right]^{2}\geq0\\
    b&=\f{\ol{P^{kk}\,\p_{1}\,\gamma_{kk}}}{\ol{\gamma_{11}}\left(1-\ve\right)}\ol{S}_{1}=\f{2\,\ol{S}_{1}}{\sqrt{\ol{\gamma_{11}}}}\sqrt{a}\\
    c&=\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}\geq0.
\end{align}
The condition for a quadratic function to be positive-definite is that $a>0$ (concave-up) and $b^{2}-4\,a\,c<0$ (no real roots). The first condition is clearly met. For the second condition:
\begin{equation}
    \f{4\left(\ol{S}_{1}\right)^{2}}{\ol{\gamma_{11}}}a-4\,a\left(\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}\right)=4\,\ol{S}_{1}\,\ol{S}^{1}\,a-4\,a\ol{D}^{2}-4\,a\,\ol{S}_{1}\,\ol{S}^{1}=-4\,a\,\ol{D}^{2}<0.
\end{equation}
So, the square root is always positive.

>>>>

\subsubsection{Time-Step Derivation/CFL Condition}
For the first-order DG method using forward-Euler time-stepping, we evolve the vector of conserved variables as:
\begin{equation}\label{Eq:1stOrderDG}
    \ol{\bU}^{n+1}_{i}=\ol{\bU}^{n}_{i}-\eta_{i}\left[\hat{\bF}\left(\ol{\bU}^{n}_{i},\ol{\bU}^{n}_{i+1}\right)-\hat{\bF}\left(\ol{\bU}^{n}_{i-1},\ol{\bU}^{n}_{i}\right)\right],
\end{equation}
where
\begin{equation}
    \ol{\bU}_{i}\equiv\f{1}{\Delta x_{i}}\int_{x_{\imh}}^{x_{\iph}}\bU_{i}\,dx,
\end{equation}
$\eta_{i}\equiv\Delta t_{i}/\Delta x_{i}$, and $\hat{\bF}$ is the numerical flux. In this document we use the local Lax-Friedrichs flux, defined as:
\begin{equation}\label{Eq:LLF}
    \hat{\bF}\left(a,b\right)=\f{1}{2}\left[\bF\left(a\right)+\bF\left(b\right) - \alpha_{ab}\left(b-a\right)\right],
\end{equation}
where $a$ and $b$ represent the state of the fluid in two different elements, $\alpha_{ab}$ is an estimate for the wave-speed:
\begin{equation}
    \alpha_{ab}=\text{max}\left[\alpha\left(a\right),\alpha\left(b\right)\right],
\end{equation}
and $\alpha$ is the largest (in absolute value) eigenvalue of the flux-Jacobian:
\begin{equation}
    \alpha=\left|\left|\pderiv{\bF}{\bU}\right|\right|.
\end{equation}
Using this we define the following variables:
\begin{equation}\label{Eq:EigVals}
    \alpha_{\iph}=\text{max}\left[\alpha\left(\ol{\bU}_{i}\right),\alpha\left(\ol{\bU}_{i+1}\right)\right],\hspace{3em} \alpha_{\imh}=\text{max}\left[\alpha\left(\ol{\bU}_{i-1}\right),\alpha\left(\ol{\bU}_{i}\right)\right].
\end{equation}

@@

\subsubsection{Set of Admissible States}
We again consider a one-dimensional system of conservation laws, but this time with a curvilinear metric:
\begin{equation}
    \pd{\left(\sqrtgm\,\bU\right)}{t}+\pd{\left(\sqrtgm\,\bF\right)}{1}=\sqrtgm\,\bQ,
\end{equation}
where $\bU$ is given by:
\begin{equation}
    \bU\longrightarrow\begin{pmatrix} D\\ S_{1}\\ \tau\end{pmatrix}=\begin{pmatrix} \rho\,W\\ \rho\,h\,W^{2}\,v_{1}\\ \rho\,W\left(h\,W-1\right)-p\end{pmatrix}=\begin{pmatrix} \rho\,W\\ \rho\,h\,W^{2}\,\gamma_{1j}\,v^{j}\\ \rho\,W\left(h\,W-1\right)-p\end{pmatrix},
\end{equation}
$\bF\left(\bU\right)$ are the fluxes of those conserved quantities:
\begin{equation}
    \bF\left(\bU\right)\longrightarrow\begin{pmatrix}D\,v^{1}\\ S^{1}\,v_{1}+p\,\delta^{1}_{~1}\\ S^{1}-D\,v^{1}\end{pmatrix}=\begin{pmatrix}\rho\,W\,v^{1}\\ \rho\,h\,W^{2}\,v^{1}\,v_{1}+p\\\rho\,h\,W^{2}\,v^{1}-D\,v^{1}\end{pmatrix}=\begin{pmatrix}\rho\,W\,v^{1}\\ \rho\,h\,W^{2}\,\gamma_{1j}\,v^{1}\,v^{j}+p\\\rho\,h\,W^{2}\,v^{1}-D\,v^{1}\end{pmatrix},
\end{equation}
and $\bQ$ is a source term:
\begin{align}
    \bQ\longrightarrow\begin{pmatrix}0\\\f{1}{2}\,P^{kj}\,\pd{\gamma_{kj}}{1}\\0\end{pmatrix}&=\begin{pmatrix}0\\ \f{1}{2}\left[P^{11}\,\pd{\gamma_{11}}{1}+P^{22}\,\pd{\gamma_{22}}{1}+P^{33}\,\pd{\gamma_{33}}{1}\right] \\0\end{pmatrix}\\
    &=\begin{pmatrix}0\\ P^{11}\,h_{1}\,\pd{\,h_{1}}{1}+P^{22}\,h_{2}\,\pd{\,h_{2}}{1}+P^{33}\,h_{3}\,\pd{\,h_{3}}{1} \\0\end{pmatrix},
\end{align}
where we have used the fact that $\gamma_{jj}=\left(h_{j}\right)^{2}$. The $P^{kj}$ are components of the pressure tensor:
\begin{equation}
    P^{kj}=S^{k}\,v^{j}+p\,\gamma^{kj}=\gamma^{kj}\left[S_{a}\,v^{a}+p\right].
\end{equation}

@@

It is clear that the first requirement for $\bH_{2}$ is met, i.e. $\ol{D}>0$. The second requirement is:
\begin{align}
\ol{D}+\ol{\tau}&>\sqrt{\ol{D}^{2}+\left[\ol{S}_{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}\right] \left[\ol{S}^{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\p^{1}\,\gamma_{kk}}\right] }\\
&=\sqrt{\ol{D}^{2}+\left[\ol{S}_{1}+\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}\right] \left[\ol{S}^{1}+\gamma^{11}\f{\Delta t}{2\left(1-\ve\right)}\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}\right] }\\
&=\sqrt{\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}+\gamma^{11}\left[\f{\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}}{2\left(1-\ve\right)}\right]^{2}\left(\Delta t\right)^{2}+\f{\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}}{2\left(1-\ve\right)}\left[\ol{S}_{1}\,\gamma^{11}+\ol{S}^{1}\right]\Delta t}\\
&=\sqrt{a\left(\Delta t\right)^{2}+b\,\Delta t+c}\geq0,
\end{align}
where
\begin{align}
    a&=\left[\f{\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}}{2\,\sqrt{\ol{\gamma_{11}}}\left(1-\ve\right)}\right]^{2}\geq0\\
    b&=\f{\ol{P^{kk}\,\pd{\gamma_{kk}}{1}}}{\ol{\gamma_{11}}\left(1-\ve\right)}\ol{S}_{1}=\f{2\,\ol{S}_{1}}{\sqrt{\ol{\gamma_{11}}}}\sqrt{a}\\
    c&=\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}\geq0.
\end{align}
The condition for a quadratic function to be positive-definite is that $a>0$ (concave-up) and $b^{2}-4\,a\,c<0$ (no real roots). The first condition is clearly met. For the second condition:
\begin{equation}
    \f{4\left(\ol{S}_{1}\right)^{2}}{\ol{\gamma_{11}}}a-4\,a\left(\ol{D}^{2}+\ol{S}_{1}\,\ol{S}^{1}\right)=4\,\ol{S}_{1}\,\ol{S}^{1}\,a-4\,a\ol{D}^{2}-4\,a\,\ol{S}_{1}\,\ol{S}^{1}=-4\,a\,\ol{D}^{2}<0.
\end{equation}
So, the square root is always positive.
